\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-21 chapters/bci.tex
Add gaze disabilities

\todo{Shortly describe challenges from the patient point of view, related to
this thesis}

@@

@@

Finally, the loop can be closed by coupling this output can be coupled to a
device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact
with their environment.
On the other hand, the actions performed by the \ac{bci} can themselves influence
the user's brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of
Parkinson's disease, or the cochlear implant as an auditive prosthesis.
More indirectly modulation can be achieved by sensory stimulation.
This can involve specific, paradigm-related sensory input, like haptic feedback when a
movement \ac{bci} detects an intended action, or simply presenting selected actions
back to the user.
The user's brain will then adapt through reinforcement learning, causing
changes in behavior or strategy.
Due to neuroplasticity, the brain's ability to adapt, this can have a positive
impact on \ac{bci} performance and opens the pathway to rehabilitation applications.

@@

\todo{why is spatial resolution good}

@@

Larger implanted electrodes are referred to as intracranial ac{eeg} (iac{eeg}).
Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
and closed-loop adaptive control for Deep Brain Stimulation to mitigate
Parkinson's disease symptoms~\cite{Arlotti2018}.
Since these still penetrate the cortex, electrodes that sit on top of the
cortex are more popular, as they are considered less invasive.
Electrocorticography (ECoG) is a powerful method
with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019} with a
higher number of smaller electrodes per surface area allow for more precise
measurements.

Some impressive recent breakthroughs\todo{cite} have been realized using
invasive mictroelectrode arrays and µECoG.
Recent advances in recording technology focus on improving implant durability and
lifetime\todo{cite flexible}, or on balancing the invasiveness tradeoff by
finding new, minimally invasive ways to record from closer to the cortex.
The Synchron Stentrode~\cite{Mitchell2023}, for instance, can be delivered
through the bloodstream via a catheter, removing the need for open brain
surgery.
However

ac{eeg} cheap and practical, non-invasive, high-temporal resolution, but noise, suffers from volume
conduction hence low spatial resolution
what does ac{eeg} measure?

different recording technologies and level of invasiveness + infographic
cost, ease of use
\section{\ac{bci} Paradigms}
\label{sec:bci-paradigms}
\fullpagefig{figures/bci/paradigms.eps}{caption}{fig:bci-paradigms}
\todo{cite A comprehensive review of ac{eeg}-based brain–computer interface
paradigms}
passive:
also monitoring for diagnostic purposes
emotion recognition
neurofeedback

>>>>

Independent home use of a brain-computer interface by people with
amyotrophic lateral sclerosis.
\todo{Shortly describe challenges from the patient point of view, related to
this thesis}
\todo{do something with Wolpaw (2018) Independent home use of a brain-computer interface by people with
amyotrophic lateral sclerosis.}

@@

\todo{put focus on communication BCI}

@@

Finally, the loop can be closed by coupling this output can be coupled to a
device or actuator.
There are two aspects to this.
On the one hand, the \ac{bci} gains its function by allowing the user to interact
with their environment.\todo{put focus on communication BCI}
On the other hand, the actions performed by the \ac{bci} can themselves influence
the user's brain activity, creating an adaptive system.
A direct form of closed-loop \ac{bci} is the use of a neurostimulator.
Examples of this are deep brain stimulation to mitigate the symptoms of
Parkinson's disease, or the cochlear implant as an auditive prosthesis.
More indirectly modulation can be achieved by sensory stimulation.
This can involve specific, paradigm-related sensory input, like haptic feedback when a
movement \ac{bci} detects an intended action, or simply presenting selected actions
back to the user.
The user's brain will then adapt through reinforcement learning, causing
changes in behavior or strategy.
Due to neuroplasticity, the brain's ability to adapt, this can have a positive
impact on \ac{bci} performance and opens the pathway to rehabilitation applications.

@@

\todo{why is spatial resolution a good thing}
\todo{explain with each method what it actually is (e.g. electrodes placed
at...)}

@@

Larger implanted electrodes are referred to as intracranial ac{eeg} (iac{eeg}).
Depth electrodes such as those used in stereo-ac{eeg} \ac{bci}~\cite{Wu2024} tasks
and closed-loop adaptive control for Deep Brain Stimulation to mitigate
Parkinson's disease symptoms~\cite{Arlotti2018}.
Since these still penetrate the cortex, electrodes that sit on top of the
cortex are more popular, as they are considered less invasive.
\ac{ecog} is a powerful method
with many applications in \ac{bci}~\cite{Schalk2011} and epilepsy diagnosis.
Newer developments like micro-ECoG (µECoG)~\cite{Shokoueinejad2019} with a
higher number of smaller electrodes per surface area allow for more precise
measurements.

Some impressive recent breakthroughs in speech and motor \ac{bci} for communication
\todo{cite} have been realized using
invasive mictroelectrode arrays~\cite{Willett2021} and high-density
\Ac{ecog}\cite{Metzger2023}.
Furthermore, recent advances in recording technology focus on improving implant
durability and resolution\todo{cite flexible}, and on balancing the invasiveness
tradeoff by finding new, minimally invasive ways to record from closer to the cortex.
The Synchron Stentrode~\cite{Mitchell2023}, for instance, can be delivered
through the bloodstream via a catheter, removing the need for open brain
surgery.
However, surveys~\cite{Huggins2011, Huggins2015, Branco2021} in different
potential communication \ac{bci} user groups consistently show that non-invasive
recording methods are preferred over implanted electrodes, unless the added value of invasive BCI is
sufficiently large~\cite{Kageyama2020}.\todo{due to what reasons}
This provides justification for focusing part of the current research effort on
non-invasive BCI.

\Ac{eeg} is a de facto non-invasive electrophysiology standard, in clinical
neurology practice as well as in \ac{bci} research.
Developed in 1924, it is cheap, relatively practical and can offers a
temporal resolution.

outside skull, environmental/electrical noise, volume conduction activity from
everywhere in the brain.

cheap and practical, non-invasive, high-temporal resolution, but noise, suffers from volume
conduction hence low spatial resolution
what does ac{eeg} measure?

cost, ease of use

\section{\ac{bci} Paradigms}
\label{sec:bci-paradigms}
\fullpagefig{figures/bci/paradigms.eps}{caption}{fig:bci-paradigms}
\todo{cite A comprehensive review of ac{eeg}-based brain–computer interface
paradigms}
passive:
also monitoring for diagnostic purposes
emotion recognition
neurofeedback

### 2024-09-21 chapters/gaze_independence.tex
Recording technologies
Add gaze disabilities

One of the goals of Brain-computer interfacing is establishing a communication
channel that does not rely on speech or
muscular activity~\cite{Naci2012,Chaudhary2016}, which in turn can provide
solutions to paralyzed individuals.
In the strictest interpretation, this also means that an interface should not
rely on the control of eye muscles which allow the subject to redirect their
gaze or stare at a target.

In fact, many patients in the BCI target population can suffer to some extent
from eye motor disability, requiring BCIs adapted to their condition.
Table~\ref{tab:incidence} reports the relatively high frequencies of
eye motor problems, which can range from minor (nystagmus, other tremors, gaze
fatigue or discomfort, \ldots), to severe (partial ophtalmoplegia, involuntary
movements, impaired pursuit, \ldots) and even complete ophtalmoplegia.
issues and even to partial or complete ophtalmoplegia.
\begin{table}[h]
  \sffamily
  \footnotesize
	\centering
	\begin{tabular}{@{}l|rrrrr|r@{}}
              & \bfseries ALS  & \bfseries MS   & \bfseries Stroke  &\bfseries DMD &\bfseries SMA  &\bfseries LiS  \\ \hline
		\bfseries Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
		\bfseries Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
		\bfseries Complete & 17\% & -    & +       & - & - & 2\%  \\
	\end{tabular}
	\caption{Incidence of eye motor impairment in selected BCI target
		patient populations. (ALS: Amyotrophic Lateral Sclerosis, MS: Multiple Sclerosis,
		DMD: Duchenne's Muscular Dystrophy, SMA: Spinal Muscular Atrophy, LiS:
		Locked-In-Syndrome).}\label{tab:incidence}
\end{table}
Among the most affected patients are ALS patients, especially those suffering
from bulbar onset ALS~\cite{Guo2022}.
Stroke, brain-stem and cerebellar stroke.
Furthermore, patients with Locked-in Syndrome
\footnote{Multiple definitions of Locked-in Syndrome are encountered in
BCI and neurological literature. Some definitions include only
tetraplegic patients without eye movements used for communications. Others distinguish Complete
Locked-in Syndrome (CLIS) with full body paralysis, including no eye motor
control at all, from a LiS state with some preserved eye movements or minor
motor output.
While some definitions only include patients who suffered stroke or traumatic
brain injury to the brain stem, it can also generally refer to the state of full body paralysis
or loss of muscle tone, combined with the inability to speak, such as occurs in
late stage ALS.
In clinical reality, every tetraplegic patient has their
unique set of preserved capabilities.}
\todo{fill out which definition we use, wolpaw criteria}
\todo{cite studies per pathology}
Gaze-independent paradigms are particularly
promising for individuals with impaired eye motor control, such as those
suffering from specific types or stages of Amyotrophic Lateral Sclerosis (ALS),
Multiple Sclerosis (MS), (brain stem) stroke or Friedreich's Ataxia (FA)

@@

\section{Gaze-independent BCI paradigms}
\label{sec:gaze-independence-paradigms}
Gaze-independent ERP-based BCIs~\cite{Riccio2012, Aloise2012} can be realized in three
ways. Firstly, non-visual stimulation paradigms such as auditory and somatosensory paradigms
do not rely on gaze redirection but often result in lower information transfer
rates, increased mental effort and user-dependent variability~\cite{Reichert2020b}.
\todo{insert table from riccio in text as numbers, not as table}

@@

\subsection{Approach}

\subsubsection{Interface design}

@@

\subsubsection{Decoders}

@@

\subsubsection{Experiments}
\todo{Healthy participant study}
We designed a visual oddball interface that helps us study and adapt
to the effects of eye motor impairment on the ERP, when using existing state-of-the-art
decoders and on our own proposed decoders.
In a first series of experiments, we recorded data with this interface from healthy participants.
The goal of these experiments is to benchmark
gaze-independent ERP decoding algorithms.
We denote this dataset as the Covert Visuospatial Attention ERP dataset
(CVSA-ERP)
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).

>>>>

One of the goals of Brain-computer interfacing is establishing a communication
channel that does not rely on speech or muscular activity~\cite{Naci2012,Chaudhary2016},
which in turn can provide solutions to paralyzed individuals.
In the strictest interpretation, this also means that an interface should not
rely on the control of eye muscles used to redirect the gaze or for blinking.
\todo{patients suffering paralysis due to damage to the CNS (stroke, spinal
chord injury, tbi) or neurodegenerative diseases}

In fact, many patients in the BCI target population can suffer to some extent
from eye motor disability, requiring BCIs adapted to their condition.
Table~\ref{tab:incidence} reports the relatively high frequencies of
eye motor problems, which can range from minor (nystagmus
\footnote{
Involuntary, rhythmic, and repetitive eye movements recognizable by their
consistent directionality (horizontal, vertical, or rotational).
}, other eye tremors
\footnote{
These can include square-wave jerks, saccadic intrusions, microtremors or
microsaccades while resting or fixating the gaze.
}, gaze fixation fatigue or discomfort, \ldots), to severe (partial ophtalmoplegia
\footnote{
Weakness or limited paralysis of one or more of the muscles that control eye
movement, leading to restricted eye motion, but not complete paralysis.
Often, a specific movement direction (up-down, left-right) is preserved.
}, involuntary movements, impaired pursuit, \ldots) and even complete
ophtalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
This not only has an effect on the patients' vision and coordination, but also
on their ability to operate a visual \ac{bci}.

\begin{table}[h]
  \sffamily
  \footnotesize
	\centering
	\begin{tabular}{@{}l|rrrrr|r@{}}
              & \bfseries ALS  & \bfseries MS   & \bfseries Stroke  &\bfseries DMD &\bfseries SMA  &\bfseries LiS  \\ \hline
		\bfseries Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
		\bfseries Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
		\bfseries Complete & 17\% & -    & +       & - & - & 2\%  \\
	\end{tabular}
	\caption{Incidence of eye motor impairment in selected BCI target
		patient populations. (ALS: Amyotrophic Lateral Sclerosis, MS: Multiple Sclerosis,
		DMD: Duchenne's Muscular Dystrophy, SMA: Spinal Muscular Atrophy, LiS:
		Locked-In-Syndrome). $+$: frequent, $-$: infrequent.}\label{tab:incidence}
\end{table}
\todo{include FA}

Among the most affected patients are stroke patients~\cite{Pollock2011}, mostly
due to brain-stem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
\todo{summarize and cite Rowe2019}
These patients can suffer from severe or complete eye motor impairment immediately
immediately from the onset of their condition.
In \ac{als}, a progressive disease, oculomotor issues are also fairly common.
Although eye movement is often cited as one of the longest preserved
capabilities in \ac{als}, \textcite{Guo2022} show that minor issues are still
fairly common,
especially for patients with bulbar onset \ac{als}.
Furthermore \textcite{Hayashi1991} show that, as these patients are supported to beyond
respiratory failure, the disease progression will eventually also involve
paralysis of the eye muscles.
Various forms of eye movement abnormalities also occur often and are especially well studied in
\ac{ms}~\cite{Mueri1985,Prasad2010,Castelnovo2016,Serra2018,Polet2020}.
These abnormalities can be minor or severe, seldom progressing to complete
paralysis.
However, \ac{ms} often comes with vision loss, further complicating interaction
with visual \acp{bci}
\todo{FA}
Patients from other neurodegenerative diseases like
\ac{sma}~\cite{Anagnostou2021} and \ac{dmd}~\cite{Lui2001} are
sometimes also interested in \ac{bci} use, but their eye motor capabilities
mostly preserved.

\newcommand\fnlis{\footnote{
Multiple definitions of \ac{lis} are encountered in
BCI and neurological literature.
Some definitions include only tetraplegic patients without eye movements used
for communications.
Others distinguish Complete Locked-in Syndrome (CLIS) with full body paralysis,
including no eye motor control at all, from a \ac{lis} state with some preserved eye
movements or minor motor output.
While some definitions only include patients who suffered stroke or traumatic
brain injury to specific regions in the brain (midbrain, brainstem,
cerebellum),
it can also generally refer to the state of full body paralysis
or loss of muscle tone incurred in neurodegenerative diseases, combined with
the inability to speak, such as occurs in late stage ALS.
In clinical reality, every tetraplegic patient has their own unique set of
preserved capabilities and from an solution-oriented \ac{bci} engineering point
of view, the etiology of the symptoms can be abstracted away.
}}

Ultimately, patients with \ac{lis} from one of the main
\ac{bci}.
In this work, we'll define \ac{lis} patients as those a situation of (near)
complete paralysis and difficulties or the inability to communicate without
assistive technology, combined with severe or complete eye motor
impairment\fnlis.
\todo{from a solution engineering point of view, the etiology does not matter that
much}.
\todo{fill out which definition we use, wolpaw criteria}
\todo{cite studies per pathology}
\todo{check definition in Wolpaw\&Wolpbaw Brain-Computer Interfaces: Principles
and Practice}
\Ac{lis} patients are are (nearly) completely paralyzed and have
effort communicating without assistive technology.

@@

\section{Gaze-independent BCI paradigms}
\label{sec:gaze-independence-paradigms}
Gaze-independent ERP-based BCIs~\cite{Riccio2012, Aloise2012} can be realized in three
ways. Firstly, non-visual stimulation paradigms such as auditory and somatosensory paradigms
do not rely on gaze redirection but often result in lower information transfer
rates, increased mental effort and user-dependent variability~\cite{Reichert2020b}.
\todo{insert table from riccio in text as numbers, not as table}
\todo{true for non-invasive}

@@

\section{Approach}

\subsection{Interface design}

@@

\subsection{Decoders}

@@

\subsection{Experiments}
\todo{Healthy participant study}
We designed a visual oddball interface that helps us study and adapt
to the effects of eye motor impairment on the ERP, when using existing state-of-the-art
decoders and on our own proposed decoders.
In a first series of experiments, we recorded data with this interface from healthy participants.
The goal of these experiments is to benchmark
gaze-independent ERP decoding algorithms.
We denote this dataset as the Covert Visuospatial Attention ERP dataset
(CVSA-ERP)
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).
